---
title: "Tidyverse3"
author: "Thomas R O'Neil"
date: "2025-03 (Updated: `r Sys.Date()`)"
output: 
  html_document:
    code_folding: hide
    theme: flatly
    css: https://drthomasoneil.github.io/CVR-site/assets/.style.css
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=F, warning=F, error=F}
library(tidyverse)
library(knitr)
library(ggpubr)

theme_set(theme_classic())

set.seed(1337)
```

# Introduction

Add notes on the experiment... 

# Methods

```{r, include=F}
samples = 3
stains = 6
```

Across `r samples` samples, I had to create a master mix for `r stains` stains.

```{r, eval=F}
dir.create("raw")
download.file(url="https://raw.githubusercontent.com/DrThomasOneil/analysis-user-group/refs/heads/main/docs/r-tutorial/assets/synthetic_data.csv", 
              destfile = "raw/data.csv", method='curl', mode='wb')
download.file(url="https://raw.githubusercontent.com/DrThomasOneil/analysis-user-group/refs/heads/main/docs/r-tutorial/assets/meta.csv", 
              destfile = "raw/meta.csv", method='curl', mode='wb')
```

```{r}
data <- read.csv("raw/data.csv")
meta <- read.csv("raw/meta.csv")
```

<details><summary>Panel</summary>

```{r}
knitr::kable(table(meta$clinical), align = 'cc')
```

</details>

# Analysis{.tabset .tabset-fade}

## Summary of Analysis

- QC

- Visualise data

- Statistics

- Conclusions

## QC and tidy up

```{r QC, eval=F}
table(data$experiment)
table(data$donor)
table(data$tissue) # issue found
table(data$layer) 
table(data$group) # issue found
```

```{r QC2}
data$tissue[grepl("abdo", data$tissue)] <- "Abdomen"
data$tissue[grepl("Abdo$", data$tissue)] <- "Abdomen"
```


## Visualisations{.tabset .tabset-fade}


### CD4:CD8 ratio

CD4 ratio to CD8 in Epithelium and Underlying mucosa of different tissues.

```{r CD4CD8ratio, fig.width=5, fig.height=4}
data %>%
  mutate(ratio = CD4/CD8) %>%
  ggplot(aes(tissue, y=ratio, fill=layer))+
  geom_boxplot(show.legend = F)+
  labs(title="Ratio of CD4:CD8", y="CD4:CD8", x="")+
  theme(axis.text.x = element_text(size=12))
```

### Percent CCR5 of CD4

sfghdfg

```{r CCR5%, fig.width=5, fig.height=4, eval=F}
data %>%
  mutate(percent = 100*CCR5/CD4) %>%
  ggplot(aes(tissue, y=percent, fill=layer))+
  geom_boxplot(show.legend = F)+
  labs(title="CCR5% of CD4", y="CCR5%", x="")+
  theme(axis.text.x = element_text(size=12))

```

## Statistics

Compare CCR5 % in epithelium of labia and vagina

```{r}
vagina <- data %>%
  mutate(CCR5_per = 100*CCR5/CD4) %>%
  filter(tissue == "Vagina" & layer == "Epithelium") %>% 
  select(CCR5_per) %>% 
  unlist()

labia <- data %>%
  mutate(CCR5_per = 100*CCR5/CD4) %>%
  filter(tissue == "Labia" & layer == "Epithelium") %>% 
  select(CCR5_per) %>% 
  unlist()

results = t.test(vagina, labia)

print(paste0("The p value for CCR5% of Vagina and labia epithelium is ", results$p.value))
```

```{r}
p0 = data %>% 
    mutate(percent = 100*CD4/CD3) %>% 
  filter(tissue %in% c("Labia", "Vagina") & layer == "Epithelium") %>% 
  ggplot(aes(tissue, percent))+geom_boxplot()+geom_jitter()+
  stat_compare_means(
    method="t.test", 
    comparisons = list(c("Labia", "Vagina")),
    label = "p.signif"
  )+theme(axis.text.x = element_text(size=15, angle=45, hjust=1), 
        axis.text.y = element_text(size=20), 
        )+
  labs(title="A)", y="Percent CD4", x="")

p1 = data %>% 
    mutate(percent = CD4/CD8) %>% 
  filter(tissue %in% c("Labia", "Vagina") & layer == "Epithelium") %>% 
  ggplot(aes(tissue, percent))+geom_boxplot()+geom_jitter()+
  stat_compare_means(
    method="t.test", 
    comparisons = list(c("Labia", "Vagina")),
    label = "p.signif"
  )+theme(axis.text.x = element_text(size=15, angle=45, hjust=1), 
        axis.text.y = element_text(size=20), 
        )+
  labs(title="B)", y="Ratio", x="")

p2 = data %>% 
    mutate(percent = 100*CCR5/CD4) %>% 
  filter(tissue %in% c("Labia", "Vagina") & layer == "Epithelium") %>% 
  ggplot(aes(tissue, percent))+geom_boxplot()+geom_jitter()+
  stat_compare_means(
    method="t.test", 
    comparisons = list(c("Labia", "Vagina")),
    label = "p.signif"
  )+theme(axis.text.x = element_text(size=15, angle=45, hjust=1), 
        axis.text.y = element_text(size=20), 
        )+
  labs(title="C)", y="Percent CCR5", x="")

p3 = data %>% 
    mutate(percent = 100*HLADR/CD4) %>% 
  filter(tissue %in% c("Labia", "Vagina") & layer == "Epithelium") %>% 
  ggplot(aes(tissue, percent))+geom_boxplot()+geom_jitter()+
  stat_compare_means(
    method="t.test", 
    comparisons = list(c("Labia", "Vagina")),
    label = "p.signif"
  )+theme(axis.text.x = element_text(size=15, angle=45, hjust=1), 
        axis.text.y = element_text(size=20), 
        )+
  labs(title="D)", y="Percent HLADR", x="")

grid <- cowplot::plot_grid(p0,p1,p2,p3, ncol=2)

ggsave("plots/CCR5per.png",
       plot=grid, 
       height=6,
       width=6, 
       units="in", 
       dpi=600)
```


# Conclusions

### Things to remember:

**markdown**    

- `yaml`: code folding, css, theme, etc   

- additional functions:   

  - `<br>` = break    
  
  - `<hr>` = horizontal rule
  
  - If you're using my css, `<wimr>` generates the nice wimr lines. 
  
  - You can isolate parts of your markdown using `<div>...</div>` tags. These create 'containers' and are useful from an aesthetic point of view.
  
  - `{.tabset .tabset-fade}` on the end of a heading allows you wrap subheadings into tabs.

**R** 
 
- `chunks`

  - on Macs - `cmd+opt+i` creates a new chunk
  
  - `include` = to show the code chunk in the knitted file
  
  - `eval` = to evaluate the code chunk
  
  - `collapse` = to collapse the output into the code chunk, and not have two separate boxes
  
  - `echo` = whether to display output
  
  - **DO NOT** remove the set up chunk
  
  - Code chunks can be named for organisational uses. 
  
  - R can be called inline by adding something like 'r 1+1' inside back ticks
  
- **packages**

  - `tidyverse`: contains a lot of different essential packages such as tidyr, ggplot, dplyr,  and stringr
  
  - `knitr`: makes neat and tidy knitable tables
  
  - `ggpubr`: is useful for tidy ggplot images and adding statistics to figures. 


Additional Resources:

- [tidyverse](https://tidyverse.tidyverse.org/)

- [R Markdown Cheatsheet](https://rstudio.github.io/cheatsheets/rmarkdown.pdf)

- [ggplot](https://ggplot2-book.org/introduction)




 



