---
title: "Flow Cytometry Analysis"
author: "Thomas R O'Neil"
date: "2025-02 <br>(Last updated: `r Sys.Date()`)"
output: 
  html_document:
    code_folding: hide
    theme: paper
    css: https://drthomasoneil.github.io/CVR-site/assets/.style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F, warning=F)
```

```{r libraries, message=F, warning=F, }
library(tidyverse)
library(knitr)
library(ggpubr)
theme_set(theme_classic())
```

<wimr>

# Analysis{.tabset .tabset-fade}

<div>

## Plans 

- Download data and load the data in ✅ 

- QC and data cleaning

- Data Exploration

- Visualisations

- Statistics

## Download and load data

Set up folders and download data. Set eval=F to make sure this doesn't run every time. 

```{r, eval=F}
dir.create("raw")
download.file(url="https://raw.githubusercontent.com/DrThomasOneil/analysis-user-group/refs/heads/main/docs/r-tutorial/assets/synthetic_data.csv", 
              destfile = "raw/data.csv", method='curl', mode='wb')
download.file(url="https://raw.githubusercontent.com/DrThomasOneil/analysis-user-group/refs/heads/main/docs/r-tutorial/assets/meta.csv", 
              destfile = "raw/meta.csv", method='curl', mode='wb')
```

Read data in:

```{r}
data <- read.csv("raw/data.csv")
meta <- read.csv("raw/meta.csv")
```

## QC and data cleaning

Checking each column of the data - issues founds with tissue and group and corrected below.

```{r, eval=F}
table(data$experiment)
table(data$donor)
table(data$tissue) # issue found
table(data$layer) 
table(data$group) # issue found
```

### Fixing Abdo and abdomen issues

```{r, include=F}
data[ grepl("Abdo$", data$tissue),"tissue"]
```

```{r}
data[data$tissue == "abdomen","tissue"] <- "Abdomen"
data[data$tissue == "Abdo","tissue"] <- "Abdomen"
```

### Adjusting group column

Take the first letter of tissue and Epi = "E" and mucosa = "UM"
```{r}
data <- data %>%
  mutate(group = paste0(str_sub(data$tissue, 1,1), "_", ifelse(data$layer == "Epithelium", "E", "UM"))) 
```

### Checking that values aren't greater than their parents

```{r}
data %>%
  mutate(CD4CD8leftover = (100*(CD4+CD8-CD3)/CD3)) %>%
  ggplot(aes(CD4CD8leftover)) + geom_density()+geom_vline(xintercept = 10)

```

Need to check some gates and values. 

</div><wimr>

# Results{.tabset .tabset-fade}

<div>

## **Questions:**

- CD4 ratio to CD8 in E and UM of different tissues

- Ditto CD4%

- Ditto CCR5% and HLADR% and CD28% of CD4 T cells.

- Ditto CCR5 and HLADR MFI

- Should I be concerned about cell numbers - compare these metrics to CD3 count.

## CD4:CD8 and CD4 %

```{r, fig.width=8, fig.height=5}
data %>%
  mutate(ratio = CD4/CD8) %>%
  mutate(enrich = ifelse(ratio > 1, "CD4", "CD8"))%>%
  ggplot(aes(
    x=layer, 
    y=log(ratio),
    fill=layer)
    )+
  geom_boxplot(outliers=F, show.legend = F)+
  geom_jitter(aes(color= enrich), size=0.75, width=0.2)+
  geom_hline(yintercept = 0)+
  facet_wrap(~tissue)+
  scale_fill_manual(values=c("pink2", "blue2"))+
  scale_color_manual(values=c("red", "black"))+
  labs(x="", y="Log CD4/CD8", title="Ratio of CD4 to CD8 across tissues and layers", color = "Enriched in:")+
  theme(axis.text.x = element_text(angle=45, hjust=1, size=10), 
        axis.text.y = element_text(size=12))
```

Make notes...

```{r}
data %>%
  mutate(percent = 100*CD4/CD3) %>%
  ggplot(aes(
    x=layer, 
    y=percent,
    fill=layer)
    )+
  geom_boxplot(outliers=F, show.legend = F)+
  geom_jitter( size=0.75, width=0.2, show.legend = F)+
  geom_hline(yintercept = 50)+
  facet_wrap(~tissue)+
  scale_fill_manual(values=c("pink2", "blue2"))+
  labs(x="", y="% CD4", title="% CD4 across cell types")+
  theme(axis.text.x = element_text(angle=45, hjust=1, size=10), 
        axis.text.y = element_text(size=12))
```

## Percent vs total cells
```{r}
data %>%
  mutate(percent = 100*CCR5/CD4) %>%
  ggplot(aes(
    x=layer, 
    y=percent,
    fill=layer)
    )+
  geom_boxplot(outliers=F, show.legend = F)+
  geom_jitter( size=0.75, width=0.2, show.legend = F)+
  facet_wrap(~tissue)+
  scale_fill_manual(values=c("pink2", "blue2"))+
  labs(x="", y="% CCR5", title="% CCR5 of CD4+ T cells")+
  theme(axis.text.x = element_text(angle=45, hjust=1, size=10), 
        axis.text.y = element_text(size=12))+
  ylim(0,100)

data %>%
  mutate(percent = 100*CCR5/CD4) %>%
  ggplot(aes(
    x=CD4, 
    y=percent)
    )+
  geom_smooth(outliers=F, show.legend = F) + facet_wrap(~layer)
```
</div><wimr>

# Statistics

Compare CCR5% in epithelium of labia and vagina

```{r}
vagina <- data  %>% 
  mutate(percent = 100*CCR5/CD4) %>% 
  filter(tissue %in% c("Vagina") & layer == "Epithelium") %>% 
  select(percent) %>% 
  unlist()
labia <- data  %>% 
  mutate(percent = 100*CCR5/CD4) %>% 
  filter(tissue %in% c("Labia") & layer == "Epithelium") %>% 
  select(percent) %>% 
  unlist()
t.test(vagina, labia)
data %>% 
    mutate(percent = 100*CCR5/CD4) %>% 
  filter(tissue %in% c("Labia", "Vagina") & layer == "Epithelium") %>% 
  ggplot(aes(tissue, percent))+geom_boxplot()+
  stat_compare_means(
    method = "t.test", 
    label = "p.signif", 
    label.y = 100,
    tip.length = 0.05,       
    bracket.size = 0.5, 
    comparisons = list(c("Labia", "Vagina")))

```


<br><wimr><br>

<br><br><br>...

### Things to remember:

**markdown**    

- `yaml`: code folding, css, theme, etc   

- additional functions:   

  - `<br>` = break    
  
  - `<hr>` = horizontal rule
  
  - If you're using my css, `<wimr>` generates the nice wimr lines. 
  
  - You can isolate parts of your markdown using `<div>...</div>` tags. These create 'containers' and are useful from an aesthetic point of view.
  
  - `{.tabset .tabset-fade}` on the end of a heading allows you wrap subheadings into tabs.

**R** 
 
- `chunks`

  - on Macs - `cmd+opt+i` creates a new chunk
  
  - `include` = to show the code chunk in the knitted file
  
  - `eval` = to evaluate the code chunk
  
  - `collapse` = to collapse the output into the code chunk, and not have two separate boxes
  
  - `echo` = whether to display output
  
  - **DO NOT** remove the set up chunk
  
  - Code chunks can be named for organisational uses. 
  
  - R can be called inline by adding something like 'r 1+1' inside back ticks
  
- **packages**

  - `tidyverse`: contains a lot of different essential packages such as tidyr, ggplot, dplyr,  and stringr
  
  - `knitr`: makes neat and tidy knitable tables
  
  - `ggpubr`: is useful for tidy ggplot images and adding statistics to figures. 


Additional Resources:

- [tidyverse](https://tidyverse.tidyverse.org/)

- [R Markdown Cheatsheet](https://rstudio.github.io/cheatsheets/rmarkdown.pdf)

- [ggplot](https://ggplot2-book.org/introduction)








