---
title: "Collaborative Coding: Markdown, Shiny, and GitHub"
author: "Thomas O'Neil"
output:
  workflowr::wflow_html:
    toc: false
---

<!---
Notes:
  Can add your own notes here.
  ...
  
<--->

<read-time><i>**Content Coming Soon**</i></read-time><wimr>

<!--- CSS Tools for Content Creators --->
<div class="hint-info">
**Important information**
</div>
<div class="hint-success">
**Success**
</div>
<div class="hint-warning">
**Warnings**
</div>
<div class="hint-danger">
**Dangers**
</div>
<div class="hint-practice">
**Practice before moving on**
</div>
<div class="hint-terminology">
**Practice you should try before moving on**
</div>

<!--- CSS Custom lines --->
<wimr> <!--- This is a good one to separate  --->
<e>

<!--- Add this html chunk and use the CSS formatting below to add a checkbox practice list --->

```{html, include=F}
<script>
document.addEventListener('DOMContentLoaded', function () {
  const checkboxes = document.querySelectorAll('input[type="checkbox"]');
  checkboxes.forEach((checkbox) => {
    checkbox.checked = localStorage.getItem(checkbox.id) === 'true';
    checkbox.addEventListener('change', () => {
      localStorage.setItem(checkbox.id, checkbox.checked);
    });
  });
});
</script>
```

<div class="hint-practice">  
**Practice:**  
<input type="checkbox"> &nbsp;&nbsp;*General Practice Title*<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<input type="checkbox"> &nbsp;&nbsp;*TASK 1* <br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<input type="checkbox"> &nbsp;&nbsp;*TASK 2* <br>
</div><e>

<!--- Start here --->

# Introduction

Notes on the tutorial

<!--- Add this html chunk and use the CSS formatting below to add a checkbox practice list --->

<div class="hint-goals">
**Goals:**    
General Goal<br>
&nbsp;&nbsp;&#8226; Goal 1<br> 
&nbsp;&nbsp;&#8226; Goal 2<br> 

</div>

<read-time><i>**Estimated time**: X hours</i></read-time><wimr>


# Set up{.tabset .tabset-fade}

<!--- Include here any information about the set up and what users might expect --->

<!--- You might also include any terminology or complicated context the user might want to research prior to the workshop --->

Set up should specify the packages required, any instructions for irregular packages or setup, data prep and download instructions. 

Folder structure, for simplicity should just follow:
```text
Name_of_group/
├── YYYYMM_YourWorkshopName_Initials
│     └── raw (for downloading and loading in raw data files)
│     └── data (for saving rds as you go)
│     └── plots (for saving rds as you go)
├── YourWorkshopName
```

## Create a new project

Users should just have to press play here.

```{r, include=F}
workshop_title = "YYYYMM_Workshop_Name"
```

```{r, eval=F}
setwd("~/Desktop/analysis-user-group")
dir.create(paste0(workshop_title, "/plots"))
dir.create(paste0(workshop_title, "/raw"))
dir.create(paste0(workshop_title, "/data"))
```

## Download the data

Users should just have to press play here.

```{r, eval=F}
# e.g. if its just a csv
download.file(url="https://raw.githubusercontent.com/DrThomasOneil/analysis-user-group/refs/heads/main/docs/r-tutorial/assets/synthetic_data.csv", 
              destfile = paste0(workshop_title, "/data.csv"))

# If not. You'll need to make it very clear how to download from r, and have a backup incase it doesn't work. 

```


## Final notes

<!--- Not Necessary! --->



<wimr>

# Your Content

# Conclusions

# Acknowledgements

<e>




