

# Chapter 5: Data Visualisation with `ggplot2`

<div class="hint-goals">      
**Goals:**     
Create and customize visualizations using `ggplot2`.    
&nbsp;&nbsp;&#8226; Basics of `ggplot2`   
&nbsp;&nbsp;&#8226; Make a scatterplot, boxplot and column graph    
&nbsp;&nbsp;&#8226; Customise labels, themes, and colors.   
</div>

R is a powerful tool for visualisation. Base R has its own graphing functions, but it's. honestly a bit too simple and ugly for publication. ***ggplot2*** is perhaps the most commonly used package. A more detailed resource can be found [here](https://ggplot2-book.org/introduction). 

ggplot will be split across two chapters. The first will be the fundamentals for making figures, and the next chapter will have notes on detailed customisations. 

<read-time><i>**Read time**: 20 Minutes</i></read-time>

<wimr>


## Basics of ggplot2

ggplot2 is simply an expansive toolbox of graphics and functions that you may combine in infinitely unique ways to create figures for your specific needs and liking.

The design of a ggplot graphic is that it is built in layers:

1. **Start with the raw input data**: which object will you use? Which columns will be on the x, y, will be grouped, will be used for colours?   
2. **Next, define how the data will be plotted** - scatterplot, violin, column, boxplot, heatmap, density etc. 
3. **Add layers.** Add titles? Add a caption? Change the colour palette? Change the theme?

<div class="hint-terminology">    
**data**: The information you want to visualise.    
**layer** collection of geometic elements and statistical transformations   
**aesthetic attributes**: color, shape, size    
**geometric objects**: points, lines, bars (**geom** for short)   
**statistical transformation**: summary of the data (e.g. binning in a histogram) (**stat** for short)   
**mapping**: description of how the variables are mapped to aesthetic attributes    
**facet**: how to break up and display subsets of data    
**theme**: the finer points of display, like  font size and background colour   
</div>



<div class="hint-goals">    
**Example Question**    
What is the relationship between CCR5 and HLA-DR percentages in human tissues?   
</div>

### Scatterplot

Using this example we'll look at how the percentage of CCR5 and HLA-DR are correlated in this dataset. 

```{r}
setwd("~/Desktop/analysis-user-group/") 
data <- read.csv("~/Desktop/analysis-user-group/synthetic_data.csv") 
```


```{r}
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(x = percent_CCR5, y = percent_HLADR))+
  geom_point()
```

<details><summary>**You don't need to use pipes for ggplot:**</summary>

This will also work:

```{r, eval=F}

data <- data %>% mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4)
ggplot(data, aes(x = percent_CCR5, y = percent_HLADR))+
  geom_point()
```

<e>

</details>

Lets start with a few aesthetic things. I'll build these sequentially and point out what is changing each time.

<details><summary>**Change the titles and axes:**</summary>

```{r}
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(x = percent_CCR5, y = percent_HLADR))+
  geom_point() + 
  ggtitle("Add a title using ggtitle()")

data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(x = percent_CCR5, y = percent_HLADR))+
  geom_point() + 
  ggtitle("Change the axes with xlab() and ylab()")+
  xlab("CCR5 %")+ylab("HLA-DR %")

data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(x = percent_CCR5, y = percent_HLADR))+
  geom_point() + 
  labs(title = "Or use `labs()` to change all at once", x= "CCR5 %", y= "HLA-DR %")
```

<e>
</details>

<details><summary>**Change the overall layout:**</summary>

Use `theme_` to test out different styles. Here are my three most commonly used. 

```{r}
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(x = percent_CCR5, y = percent_HLADR))+
  geom_point() + 
  labs(title = "theme_classic()", x= "CCR5 %", y= "HLA-DR %")+
  theme_classic()
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(x = percent_CCR5, y = percent_HLADR))+
  geom_point() + 
  labs(title = "theme_minimal())", x= "CCR5 %", y= "HLA-DR %")+
  theme_minimal()
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(x = percent_CCR5, y = percent_HLADR))+
  geom_point() + 
  labs(title = "theme_bw()", x= "CCR5 %", y= "HLA-DR %")+
  theme_bw()
```


<e></details>

<details><summary>**Adjust the aesthetic attributes (categorical):**</summary>

Changing the aesthetic will let us pull more information from the simple black dots

```{r}
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(
    x = percent_CCR5, 
    y = percent_HLADR,
    color = tissue
  ))+
  geom_point() + 
  labs(title = "Color by tissue", x= "CCR5 %", y= "HLA-DR %")+
  theme_classic()

data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(
    x = percent_CCR5, 
    y = percent_HLADR,
    color = tissue
  ))+
  geom_point() + 
  labs(title = "Color by layer", x= "CCR5 %", y= "HLA-DR %")+
  theme_classic()

data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(
    x = percent_CCR5, 
    y = percent_HLADR,
    color = paste0(tissue, "_", layer)
  ))+
  geom_point() + 
  labs(title = "Color by tissue & layer", x= "CCR5 %", y= "HLA-DR %")+
  theme_classic()

data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(
    x = percent_CCR5, 
    y = percent_HLADR,
    color = tissue, 
    shape = layer
  ))+
  geom_point() + 
  labs(title = "Color by tissue & shape by layer", x= "CCR5 %", y= "HLA-DR %")+
  theme_classic()
```

<e></details>

<details><summary>**Adjust the aesthetic attributes (continuous):**</summary>

We could add a third dimension here. X and Y are the percentage values, but what if we coloured the points by a third continuous value? 

```{r}
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(
    x = percent_CCR5, 
    y = percent_HLADR,
    color = CCR5_MFI
  ))+
  geom_point() + 
  labs(title = "Color by CCR5 MFI", x= "CCR5 %", y= "HLA-DR %")+
  theme_classic()

data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4,
         percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(
    x = percent_CCR5, 
    y = percent_HLADR,
    color = log(CCR5_MFI)
  ))+
  geom_point() + 
  labs(title = "Color by log(CCR5 MFI)", x= "CCR5 %", y= "HLA-DR %")+
  theme_classic()
```

<e></details>

<wimr>

<div class="hint-goals">    
**Example Question**    
Which tissues are enriched for CCR5?
</div>

### Boxplot

<details><summary>**Boxplot vs Violin:**</summary>

```{r}
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4) %>%
  ggplot(aes(
    x = group, 
    y = percent_CCR5
  ))+
  geom_boxplot() + 
  labs(title = "geom_boxplot()", y= "CCR5 %", x= "Tissue x Layer")+
  theme_classic()

data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4) %>%
  ggplot(aes( 
    x = group, 
    y = percent_CCR5
  ))+
  geom_violin() + 
  labs(title = "geom_violin()", y= "CCR5 %", x= "Tissue x Layer")+
  theme_classic()
```

<e></details>

<details><summary>**Combining layers - boxplot & points:**</summary>


```{r}
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4) %>%
  ggplot(aes(
    x = group, 
    y = percent_CCR5
  ))+
  geom_boxplot() + 
  geom_point()+
  labs(title = "geom_boxplot() AND geom_point()", y= "CCR5 %", x= "Tissue x Layer")+
  theme_classic()
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4) %>%
  ggplot(aes(
    x = group, 
    y = percent_CCR5
  ))+
  geom_boxplot() + 
  geom_jitter()+
  labs(title = "geom_boxplot() AND geom_jitter()", y= "CCR5 %", x= "Tissue x Layer")+
  theme_classic()
```


```{r}
# The default jitter is width = 0.5 - we can reduce this 
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4) %>%
  ggplot(aes(
    x = group, 
    y = percent_CCR5
  ))+
  geom_boxplot() + 
  geom_jitter(width=0.1)+
  labs(title = "geom_boxplot() AND geom_jitter(width=0.1)", y= "CCR5 %", x= "Tissue x Layer")+
  theme_classic()
```

<e></details>


<details><summary>**Adjusting the aestheticss:**</summary>

When you're combining layers, like boxplot and point, you might want to change the aesthetic to each layer individually. So instead of changing the aes at the start, we can add aes inside the geom's.

```{r}
# The default jitter is width = 0.5 - we can reduce this 
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4) %>%
  ggplot(aes(
    x = group, 
    y = percent_CCR5
  ))+
  geom_boxplot(aes(fill = group)) + 
  geom_jitter(width=0.1, aes(color=age))+
  labs(title = "geom_boxplot(aes(fill=group))+geom_jitter(aes(color=age))", y= "CCR5 %", x= "Tissue x Layer")+
  theme_classic()
```

<e></details>

<wimr>

<div class="hint-goals">    
**Example Question**    
What is the ratio of CD4:CD8 per tissue and layer?
</div>

### Column/bar graph

Column/bar graphs are useful for summary graphs and/or statistical transformations (such as counts or histograms). 

<details><summary>**Column graph:**</summary>

```{r}
data %>%
  mutate(ratio = log(CD4/CD8)) %>%
  group_by(group)%>%
  summarize(meanratio = mean(ratio))%>%
  mutate(`Predominant Subset` = ifelse(meanratio<0, "CD8", "CD4"))%>%
  ggplot(aes(x=group, y=meanratio, fill=`Predominant Subset`))+
  geom_col()+
  theme_bw()
```


</details>

<details><summary>**Example of statistical transformations (Histogram/Bar graph):**</summary>

```{r}
data %>%   
  mutate(percent_CD4 = 100*CD4/CD3,
         percent_CD8 = 100*CD8/CD3) %>%
  ggplot(aes(x=percent_CD4))+
  geom_histogram()+
  theme_bw()+
  ggtitle("Histogram example")
# change the bin width to add more resolution
data %>%   
  mutate(percent_CD4 = 100*CD4/CD3,
         percent_CD8 = 100*CD8/CD3) %>%
  ggplot(aes(x=percent_CD4))+
  geom_histogram(binwidth = .5)+
  theme_bw()+
  ggtitle("Histogram example")

data %>%   
  mutate(predominant = ifelse(CD4>CD8, "CD4", "CD8")) %>%
  ggplot(aes(x=predominant))+
  geom_bar()+
  theme_bw()+
  ggtitle("Bar graph example: Number of samples that are predominantly CD4 or CD8")
```

<e></details>


## Summary

With these simple examples, you should have an idea of how to make and customise ggplots for your data. 

The next chapter will build upon these, such as changing the theme elements, customising the colours, combining and then saving your plots. 

[Next chapter →](https://drthomasoneil.github.io/analysis-user-group/r-tutorial/_book/chapter-6-data-visualisation-with-ggplot2-2.html) 


<div id="quote-box" style="text-align:center; font-style:italic; margin-top:20px; margin-bottom:20px;"></div>

