

# Chapter 6: Data Visualisation with `ggplot2` 2


<div class="hint-goals">      
**Goals:**     
Learn additional visualisation techniques.    
&nbsp;&nbsp;&#8226; Adjusting theme elements    
&nbsp;&nbsp;&#8226; Annotating the plots     
&nbsp;&nbsp;&#8226; facet_wrap() and grid_plot()    
&nbsp;&nbsp;&#8226; Savings plots   
</div>

<read-time><i>**Read time**: 15 Minutes</i></read-time>

<wimr>

## Adjusting theme elements

Everything in ggplot can be adjusted. Lets just go through a few of them to show you how its done, and you can explore them as you go. 


<details><summary>**Theme elements:**</summary>

```{r}
data %>%
  ggplot(aes(CD28, age, color = log(CCR5_MFI))) +
  geom_point()+
  theme_classic()+
  theme(axis.text = element_text(size=10),
        axis.title = element_text(size=12), 
        legend.position = "inside", legend.justification = c(.99,.99))
```
<e></details>


<details><summary>**Changing the scale:**</summary>

```{r}
data %>%
  ggplot(aes(CD28, age)) +
  geom_point()+
  theme_bw()+
  ylim(0,100)
```
<e></details>

<details><summary>**Changing the colours manually:**</summary>

```{r}
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4) %>%
  ggplot(aes(
    x = group, 
    y = percent_CCR5, 
    fill = group
  ))+
  geom_boxplot(width=0.5) + 
  geom_jitter(width=0.1, alpha=0.2)+
  labs( y= "CCR5 %", x= "Tissue x Layer")+
  scale_fill_manual(values = c("blue", "blue4", "red", "red4", "green", "green4"))
  theme_classic()
data %>% mutate(percent_HLADR = HLADR/CD4) %>%
  ggplot(aes(x=CD28, y=age, color = percent_HLADR)) +
  geom_point()+
  theme_bw()+
  scale_color_gradient(high = "#FF0000", low = "#0000FF") # you can search hex numbers, or match colours to a powerpoint presentation. You can use digital colour meter on Mac
data %>% mutate(percent_HLADR = HLADR/CD4) %>%
  ggplot(aes(x=CD28, y=age, color = percent_HLADR)) +
  geom_point()+
  theme_bw()+
  scale_color_viridis_c() # you can search through the options for these, as well as other scale_color and scale_fill options.
```
<e></details>

<wimr> 

## Annotating the plot

<details><summary>**Horizontal and Vertical lines:**</summary>

I've added vertical lines (vlines), where you can specify the xintercepts to draw lines through. 

I also added "annotate" layers where you can specify "text" or "segment" layers, with x, y and label arguments for 

```{r}
data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4) %>%
  ggplot(aes(
    x = group, 
    y = percent_CCR5
  ))+
  geom_boxplot(width=0.5) + 
  geom_jitter(width=0.1, alpha=0.2)+
  labs( y= "CCR5 %", x= "Tissue x Layer")+
  geom_vline(xintercept = c(2,4)+.5)+
  annotate("segment", 
           x=c(1:6)-.25, xend = c(1:6)+.25, 
           y=data%>% mutate(percent_CCR5 = 100*CCR5/CD4) %>% group_by(group) %>% summarise(mean = mean(percent_CCR5)) %>% select(mean) %>% unlist(), color="red", size=2)+
  annotate("text", 
           x=1:6, 
           y=data%>% mutate(percent_CCR5 = 100*CCR5/CD4) %>% group_by(group) %>% summarise(mean = mean(percent_CCR5)) %>% select(mean) %>% unlist(),
           label = data%>% mutate(percent_CCR5 = 100*CCR5/CD4)%>%group_by(group)%>%summarise(mean=mean(percent_CCR5)) %>% select(mean) %>% unlist() %>%round(),
           hjust = -2, color="red")+
    theme_classic()
```
<e></details>

<wimr> 

## facet_wrap() and Group

Facet wrap lets you split your points by a categorical value

<details><summary>**Facet_wrap():**</summary>

```{r}
data %>% mutate(percent_HLADR = 100*HLADR/CD4) %>%
  ggplot(aes(x=CD28, y=age, color = percent_HLADR)) +
  geom_point(aes(shape=layer))+
  theme_bw()+
  scale_color_gradient(high = "#FF0000", low = "#0000FF")+
  facet_wrap(~tissue)
```
<e></details>

<details><summary>**plot_grid():**</summary>

```{r}
library(cowplot)

plot1 = data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4) %>%
  ggplot(aes(
    x = group, 
    y = percent_CCR5
  ))+
  geom_boxplot(width=0.5) + 
  geom_jitter(width=0.1, alpha=0.2)+
  labs( y= "CCR5 %", x= "Tissue x Layer", title="Percent CCR5")+
  geom_vline(xintercept = c(2,4)+.5)+
  theme_classic()
plot2 = data %>%
  ggplot(aes(
    x = group, 
    y = log(CCR5_MFI)
  ))+
  geom_boxplot(width=0.5) + 
  geom_jitter(width=0.1, alpha=0.2)+
  labs( y= "log(CCR5 MFI)", x= "Tissue x Layer", title="CCR5 MFI")+
  geom_vline(xintercept = c(2,4)+.5)+
  theme_classic()

plot_grid(plot1,plot2, nrow=2) #or change the rows. Add 50 plots if you want to adn specify how you arrange them by rows and/or columns

```
<e></details>

<wimr> 

## Saving plots

You can save plots in a few ways.

<details><summary>**pdf/png/tiff/jpeg():**</summary>

Use `?png` to get the specific options, but here is a simple example of generating a multipage pdf. 

You call the function to "open" the pdf, plot your figures, and then close the pdf using `dev.off()`

```{r, eval=F}
pdf(file="CCR5_results.pdf", widht = 7, height=7) 

data %>%
  mutate(percent_CCR5 = 100*CCR5/CD4) %>%
  ggplot(aes(
    x = group, 
    y = percent_CCR5
  ))+
  geom_boxplot(width=0.5) + 
  geom_jitter(width=0.1, alpha=0.2)+
  labs( y= "CCR5 %", x= "Tissue x Layer", title="Percent CCR5")+
  geom_vline(xintercept = c(2,4)+.5)+
  theme_classic()

data %>%
  ggplot(aes(
    x = group, 
    y = log(CCR5_MFI)
  ))+
  geom_boxplot(width=0.5) + 
  geom_jitter(width=0.1, alpha=0.2)+
  labs( y= "log(CCR5 MFI)", x= "Tissue x Layer", title="CCR5 MFI")+
  geom_vline(xintercept = c(2,4)+.5)+
  theme_classic()

dev.off()
```

<e></details>

<details><summary>**ggsave:**</summary>

ggsave is quite simple and allows you to very easily control resolutions and size. 

```{r, eval=F}

ggsave("CCR5_percent.png", 
       plot1, 
       path = "C:/1_Yc/r-analysis-user-group/plots", 
       width = 7, height = 5, units = c('cm'), 
       dpi=300)

```

<e></details>

<wimr> 

## Some final hints

<div class="hint-info">      
&nbsp;&nbsp;&#8226; You can use show.legend=F inside geom layers to remove the legends    
&nbsp;&nbsp;&#8226; You can store your initial plots as `plot = ggplot(...)+geom_...`     
&nbsp;&nbsp;&nbsp;&nbsp;and then append layers onto this.     

</div>

[Next Chapter →](https://drthomasoneil.github.io/analysis-user-group/r-tutorial/_book/chapter-7-statistical-analysis.html)


<div id="quote-box" style="text-align:center; font-style:italic; margin-top:20px; margin-bottom:20px;"></div>
