---
title: "Introduction to R and analytic Programming"
date: "2024-12-23 (Last update: `r Sys.Date()`)"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
description: |
  This is a minimal example of using the bookdown package to write a book.
  The HTML output format for this example is bookdown::gitbook,
  set in the _output.yml file.
link-citations: yes
github-repo: DrThomasOneil/analysis-user-group
output: tufte_html_book
  
---
 
# Welcome! 
 
This is the sub-repository containing the tutorials and workshop notes for the introduction to R and analytic programming. 

A mock dataset will be provided, which is an example of summarised flow cytometry data, containing mock batches, patients, samples and measurements like percentage of subsets/MFIs.

This tutorials provided are contained within a  

<div class="hint-info">
<strong>Info:</strong> This is to accentuate important points.  
</div>

<div class="hint-success">
<strong>Success:</strong> This is good for showing positive actions or achievements.  
</div>  

<div class="hint-warning">
<strong>Warning:</strong> This is important information or non-critical warnings.  
</div>

<div class="hint-danger">
<strong>Danger:</strong> This highlights destructive actions or critical information.
</div>
 
```{r, eval=FALSE}
bookdown::render_book()
```

```{r, echo=F, eval=FALSE}
servr::daemon_stop(1:7)
bookdown::serve_book() 
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

<e>
