---
title: "Introduction to R and analytic Programming"
author: "Thomas O'Neil"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
description: |
  This is a minimal example of using the bookdown package to write a book.
  The HTML output format for this example is bookdown::gitbook,
  set in the _output.yml file.
link-citations: yes
github-repo: DrThomasOneil/analysis-user-group
---

# Welcome! 

This is the sub-repository containing the tutorials and workshop notes for the introduction to R and analytic programming. 

A mock dataset will be provided, which is an example of summarised flow cytometry data, containing mock batches, patients, samples and measurements like percentage of subsets/MFIs.

This tutorials provided are contained within a 


```{r, eval=FALSE}
bookdown::render_book()
```



## Preview book

As you work, you may start a local server to live preview this HTML book. This preview will update as you edit the book when you save individual .Rmd files. You can start the server in a work session by using the RStudio add-in "Preview book", or from the R console:

```{r eval=FALSE}
bookdown::serve_book()
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
