---
title: ""
output:
  workflowr::wflow_html:
    toc: no
--- 

<div style="text-align:center">

# Find a solution to your bugs!

**Curator**: Yuchen Li<br>

**[Contact Me `r "\U2709\UFE0F"`](mailto: yuli0321@uni.sydney.edu.au)** if you can't find your bug! ;) Include a screenshot, and/or copy of the script and error

 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(shiny)
library(reactable)
library(htmltools)
```


```{r, warning = F, message=FALSE}
content <- read.csv("docs/adit/0_content management.csv")
htmltools::browsable(
  table <- reactable(content,
                     defaultColDef = colDef(
                       align = "left",
                       headerStyle = list(background = "#ffcf72")
                     ),
                     # Define columns
                     columns = list(
                       Topic = colDef(name = "Topic", width = 220,
                                      filterable = TRUE), # Make it sticky when scrolling
                       Platform = colDef(name = "Platform", width = 80, # e.g. R/python
                                         filterable = TRUE),
                       Theme = colDef(name = "Theme", width = 100,
                                      filterable = TRUE),
                       Tag = colDef(name = "Tag", width = 220,
                                    filterable = TRUE),
                       Link = colDef(name = "Link",
                                     html =TRUE,
                                     # In a cell setting, initiate an R render function
                                     cell = function(value){
                                       # generates a link to a webpage by adding an 'a' html tag,
                                       tags$a(href = value, target = "_blank", class = "page-link", 
                                              ifelse(is.na(value)|| value == ""||is.null(value), "Coming soon", "View page"))
                                     },
                                     width = 100),
                       Recording = colDef(name = "Recording",
                                          html = TRUE,
                                          cell = function(v){
                                            tags$a(href = v, target = "_blank", class = "video-link", "View video")
                                          },
                                          width = 100)
                     ),
                     # Define other table characteristics
                     bordered = TRUE,
                     highlight = TRUE,
                     compact = TRUE,
                     # Default sorting by schedules
                     defaultSorted = list(Date = "asc"),
                     # Make the table searchable
                     searchable = TRUE,
                     # Show 8 rows max a time
                     defaultPageSize = 20,
                     # group the content together by themes
                     groupBy = "Topic",
                     paginateSubRows = TRUE,
                     theme = reactableTheme(
                       searchInputStyle = list(width = "30%")
                     )
  )
)
htmlwidgets::saveWidget(table, "./docs/0_archive.html", selfcontained = TRUE)
```

 </div>

#### If your machine is onboarded with WIMR...

You will face several restrictive issues during your analysis. You are required to submit requests for admin permissions to install new packages sometimes, and new software. If you have extensive issues with this, please make a complaint to the DRSN through Chair **[Thomas O'Neil](mailto: thomas.oneil@sydney.edu.au)** who will represent you, and advise you on how to proceed with analysis. 
