---
title: ""
output:
  workflowr::wflow_html:
    toc: no
--- 

<div style="text-align:center">

# Find a solution to your bugs!

**Curator**: Yuchen Li<br>

**[Contact Me `r "\U2709\UFE0F"`](mailto: yuchen.li@sydney.edu.au)** if you can't find your bug! ;) Include a screenshot, and/or copy of the script and error

<read-time><i>**Coming Soon!**</i></read-time>
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(shiny)
library(reactable)
library(htmltools)
```


```{r, warning = F, message=FALSE, eval=F, include=F}
bugs <- read.csv("docs/adit/bugs.csv")
bugs$Date <- as.character(bugs$Date)
htmltools::browsable(
  table <- reactable(bugs,
                     defaultColDef = colDef(
                       align = "left",
                       headerStyle = list(background = "#ffcf72", fontWeight = "bold")
                     ),
                     # Define columns
                     columns = list(
                       
                       Date = colDef(name = "Date", width = 80),
                       
                       Platform = colDef(name = "Platform", width = 80, # e.g. R/python
                                         filterable = TRUE),
                       Context = colDef(name = "Context", width = 200),
                       
                       Error.message = colDef(name = "Error message", width = 300),
                       
                       Solution = colDef(name = "Solution", width = 300)
                     ),
                     
                     # Define other table characteristics
                     bordered = TRUE,
                     highlight = TRUE,
                     compact = TRUE,
                     # Default sorting by schedules
                     defaultSorted = list(Date = "asc"),
                     # Make the table searchable
                     searchable = TRUE,
                     # Show 20 rows max a time
                     defaultPageSize = 20,
                     paginateSubRows = TRUE,
                     theme = reactableTheme(
                       searchInputStyle = list(width = "30%")
                     )
  )
)
htmlwidgets::saveWidget(table, "./docs/0_bugs.html", selfcontained = TRUE)
```

 </div>

#### If your machine is onboarded with WIMR...

And you're experiencing issues with R, RStudio or python, please reach out to **[Thomas O'Neil](mailto: thomas.oneil@sydney.edu.au)** for help navigating these issues.
<!---
You will face several restrictive issues during your analysis. You are required to submit requests for admin permissions to install new packages sometimes, and new software. If you have extensive issues with this, please make a complaint to the DRSN through Chair **[Thomas O'Neil](mailto: thomas.oneil@sydney.edu.au)** who will represent you, and advise you on how to proceed with analysis. 
<--->
