---
output:
  workflowr::wflow_html:
    toc: no
    code_folding: "hide"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning=FALSE, error=F)
```

## Flow Cytometry in R: A Hands-On Introduction

This beginner-friendly, hands-on workshop will guide users through a complete flow cytometry analysis workflow in R, from loading data to visualizing results. Through practical exercises, we’ll cover essential steps like generating histograms and scatterplots, extracting summary statistics, subsetting data, and performing dimension reduction and clustering—all while demonstrating the advantages of reproducible reports with RMarkdown.

Whether you're new to R or just curious about coding your analyses, this session will show you how accessible and powerful R can be for flow cytometry.

**Your Demonstrators:**

-   Thomas O'Neil\
-   Yuchen Li (Online)

**Author**: Yuchen Li, Thomas O'Neil

[Zoom link](https://uni-sydney.zoom.us/s/88109360455) (recorded)

# Introduction

In this workshop, we aim to give an overview of what is possible for flow analysis in R. We will touch on two workflows:

-   data visualisation and data collection

-   dimension reduction and clustering

We'll use a 25-colour flow cytometry dataset of innate lymphoid cells (ILC), natural killer cells (NK), and innate-like T cells in human intestinal tissues, after excluding dead cells, doublets, mast cells and basophils, hematopoietic stem cells, myeloid cells and granulocytes, and B cells.

The original data was published as is available in [OMIP-082](https://pubmed.ncbi.nlm.nih.gov/35018731/)

<details>

<summary>**Before going further**:</summary>

-   Dont forget to provide any feedback by emailing us at `drsn@wimr.org.au` or filling out the [feedback form](https://forms.office.com/r/DXdTAQXJSU)

-   Vote on the next workshop!

<iframe width="640px" height="200px" src="https://forms.office.com/r/TtLWFWysiC?embed=true" frameborder="0" marginwidth="0" marginheight="0" style="border: none; max-width:100%; max-height:100vh" allowfullscreen webkitallowfullscreen mozallowfullscreen msallowfullscreen>

</iframe>

</details>

<wimr>

# Setup {.tabset .tabset-fade}

We'll take you through steps from the beginning, installing packages and loading data.

## Preparation: packages and directories

Packages are collections of functions. Many are not in-built in the base R but they do pretty cool things, enabling generating pretty plots, performing data wrangling in an efficient way, etc. R also has a large developer community, with many community-developed packages available on platforms like [`Bioconductor`](https://www.bioconductor.org/) or `GitHub`. Many packages we use in biomedical sciences are uploaded on Bioconductor.

```{r, message=FALSE, warning=FALSE, eval=F}
# Install packages used in this workshop.
install.packages("tidyverse")
install.packages("ggridges")
install.packages("GGally")
install.packages("cowplot")
install.packages("data.table")
install.packages("umap")
install.packages("Rtsne")

# The package 'BiocManager' is required to install packages uploaded to Bioconductor, like FlowSOM and FuseSOM.
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("FuseSOM")
BiocManager::install("FlowSOM")

if(!require('remotes')) {install.packages('remotes')}
remotes::install_github(repo = "immunedynamics/spectre")
```


```{r, message=FALSE, warning=FALSE}
# Load packages
library(tidyverse)
library(ggridges)
library(cowplot)
library(Spectre)
library(FuseSOM)
library(FlowSOM)
library(GGally)
library(umap)
library(Rtsne)
library(data.table)

# Set the plot themes for the entire document.
theme_set(theme_classic())
```

Create some folders for plots and data.

```{r, message=FALSE, warning=FALSE, eval=F}
# Have a look at which directory you are working in.
getwd()

# Create new subdirectories.
dir.create("plots")
dir.create("data")

# This should download the file to the current directory that this Rmd is saved to.
download.file("https://raw.githubusercontent.com/DrThomasOneil/Digital-Research-Skills-Network/refs/heads/main/docs/assets/downloads/example_flowdata_omip082_20k.csv", destfile = "data/example_flowdata_omip082_20k.csv")
# Add `method = "curl"` in the download.file() if you are a Windows user and this line didn't work for you.
```

<details>

<summary>Setting seed is crucial for reproducibility:</summary>

It's a number of your choice, which pins down the "way" that how the random process will be performed when you re-run these codes. (Easier example: Say if you randomly take 10 numbers with a mean of 50, there will be tonnes of ways to generate the numbers. But a seed ensures that the report always returns the same set of numbers. It's the similar idea when you run other algorithms including randomness, e.g. clustering and generating a UMAP).

```{r}
set.seed(1337)
```

</details>

<wimr>

## Load flow cytometry data

The expression data is stored in a `.csv` file. We'll first load this in.

```{r}
# Read the .csv file in the working directory.
ILC <- read.csv("data/example_flowdata_omip082_20k.csv")
```

Have a look at the broad features of this dataset. In this dataset, the channel names are stored as column names. Each row represents a cell, storing expression values in each channel.

```{r, eval=F}
# Dimension: number of cells x number of channels.
dim(ILC)

# Check the column names.
colnames(ILC)
```

The column names are wordy, with markers, fluorophores, and channels. We just need the marker names to proceed. The first column 'X' is just indexing the number of rows. It's not necessary either.<br>

Remove the first column.

```{r}
# This keeps the 2-33 columns in the dataset, and store the change.
ILC <- ILC[, -1]
colnames(ILC)

# Same with it, you can just enter the column indexes from 2-33.
# ILC <-ILC[,c(2:33)]
```

Change the column names. Note: this is very specific manipulations on the names on THIS dataset. Don't directly copy and apply on your own work! It's just to demonstrate some ways to manipulate the column names.

```{r}
# Substitute everything after the first '_' with an empty "" in the column names that have the pattern "_" and something.
colnames(ILC) <- sub("_.*","",colnames(ILC)) #'_' is the exact character in the names, the '.' means 'any character', '*' means 'any of the preceding things'. 

colnames(ILC)
```

Fix up the broken names FVS700 and CD14&CD19.

```{r}
# This tells R to: find which of the column equals "FVS" among all columns in the data, then change it to FVS700.
colnames(ILC)[colnames(ILC) == "FVS"] <- "FVS700"
colnames(ILC)[colnames(ILC) == "CD14"] <- "CD14.CD19"

# Have a look
colnames(ILC)
```

<wimr>

## Time gate QC.

The data is already preprocessed in FlowJo. But if you are interested, you can view the Time gate, and filter the data based on this in R.

```{r}
ggplot(ILC, aes(x = Time, y = FSC.A))+
  geom_bin_2d(bins = 200, show.legend = FALSE)+
    scale_fill_gradientn(colours = c("blue", "green", "yellow", "red"))+
  geom_vline(xintercept = 4600000)

# Filter out the cells above the threshold.
ILC <- ILC[ILC$Time < 4600000,]
ggplot(ILC, aes(x = Time, y = FSC.A))+
  geom_bin_2d(bins = 200, show.legend = FALSE)+
    scale_fill_gradientn(colours = c("blue", "green", "yellow", "red"))
```

<wimr>

# Visualisations {.tabset .tabset-fade}

## Transformations

Raw Flow Cytometry data looks bad!

```{r}
ILC %>% ggplot(aes(x=CD4, y=CD3))+        
  geom_point()
```

So we transform the data to make it look better. In FlowJo, when you adjust the axes, you can see several transformations available. One is arcsinh (the inverse of hyperbolic sin, if you want to get mathy). Smaller values centre around zero, and larger values separate, giving the appearance that we're familiar with.

We are also able to adjust the level of transformation by dividing the values by some cofactor.

```{r, fig.height=7, fig.width=7}
# decide on a cofactor - try changing it and seeing how the data changes
cofactor = 500
ILC_transf <- asinh(ILC[,7:31]/cofactor) #7:31 are the columns with markers.

ILC_transf <- cbind(ILC_transf, ILC[,c(1:6,32)]) # Add the first 6 columns & time back in.

ILC_transf %>% ggplot(aes(x=CD4, y=CD3))+
  geom_bin_2d(bins = 200, show.legend=T)
```

You can view it for all markers together.

```{r, fig.height=10, fig.width=10, warning=FALSE}
# Initialize an empty list to store the plots
plots <- list()
# Define the x-axis marker for the plots
xaxis = "CD39" # Choose an x-axis marker for comparison

# Loop through each column in the dataset
for(i in 1:ncol(ILC_transf)) {
  # Get the marker name for the current column
  Marker <- colnames(ILC_transf)[i]
  
  # Generate a 2D bin plot for the current marker against the chosen x-axis marker
  p <- ILC_transf %>% 
        ggplot(aes_string(x = xaxis, y = Marker)) +        
        geom_bin_2d(bins = 200, show.legend = FALSE) + # Create a heatmap-style scatter plot
        scale_fill_gradientn(colors = c("blue", "yellow", "red")) # Define the color gradient
  
  # Add the generated plot to the list
  plots[[i]] <- p
}

# Combine all the plots into a grid layout and display them
cowplot::plot_grid(plotlist = plots)
rm(plots)
```

This code demonstrates how you would manually transform individual columns in the dataset using different cofactors for each marker. It creates a new dataset (`ILC_transf_indiv`) where specific columns are transformed using the `asinh` function with their respective cofactors. This allows for more granular control over transformations.

```{r, eval=F}
ILC_transf_indiv <- ILC
ILC_transf_indiv[, "CD3"] <- asinh(ILC[,"CD3"]/1000)
ILC_transf_indiv[, "CD4"] <- asinh(ILC[,"CD4"]/500)
# and so on...

# Visualize it
ILC_transf_indiv %>% ggplot(aes(x=CD3, y=CD4))+        
        geom_bin_2d(bins = 200, show.legend=F) +
        scale_fill_gradientn(colors = c("blue",  "yellow", "red"))

rm(ILC_transf_indiv)
```

<wimr>

## Histograms

We can visualize the marker expressions via histograms and density plots.

```{r}
# For example, this generates a histogram of CD3 expression.
ILC_transf %>%
  ggplot(aes(x = CD3))+
  # Plot the histogram, and change the fill colour to lightblue, the line colour to black.
  geom_histogram(bins = 35, fill = "lightblue", colour ="black")+
  # Add labels
  ylab("Frequency")

# A density plot.
ILC_transf %>%
  ggplot(aes(x = CD3))+
  geom_density(linewidth = 0.7, colour = "red4", aes(fill = "red"), alpha = 0.5,show.legend = FALSE)+
  ylab("Density")

```

We can loop through the channels and create a density plot for all 25 markers.

```{r, message = FALSE}
# Initialise an empty list to store the plots
plots<- list()

# Loop through each column in the dataset
for (i in 1:ncol(ILC_transf)){

  # Get the marker name for the current column
  Marker <- colnames(ILC_transf)[i]

  # Generate the histogram for the current marker
  p <- ILC_transf %>%
  ggplot(aes(x = .data[[Marker]]))+
    geom_density(linewidth = 0.7, colour = "red4", aes(fill = "red"), alpha = 0.5,show.legend = FALSE)+
    labs(x=Marker,y="")+
    theme(axis.text = element_text(size=1))

  # Add the generated plot to the list
  plots[[i]] <- p
}

plot_grid(plotlist = plots[1:25])
rm(plots)
```

<wimr>

## Scatter plot

Like what we do in Flowjo, we can visualize 2 markers' expressions, We can generate scatterplots.

```{r}
ILC_transf %>%
  ggplot(aes(x = CD4, y = CD3))+
  geom_point(color="blue", alpha=0.1)

# Generate a scatterplot with a density overlay.
ILC_transf %>% ggplot(aes(x=CD4, y=CD3))+        
  geom_bin_2d(bins = 200, show.legend = F)+
  scale_fill_gradientn(colors = c("blue",  "yellow", "red"))

```

<wimr>

## Contour plot

And a 2-marker countour plot. You can easily make it more stylish by adding layers in ggplot!

```{r collapse=T}

# A contour plot of CD3 and CD4 expression values.
ILC_transf %>%
  ggplot(aes(x = CD3, y = CD4))+
  geom_density_2d(colour = "black", bins=50)+
  xlab("CD3 BUV563")+
  ylab("CD4 APC-Cy7")

# A contour plot of CD3 and CD4 expression over a scatter plot.
ILC_transf %>%
  ggplot(aes(x = CD3, y = CD4))+        
  geom_point(color = "grey", alpha=0.5)+
  geom_density_2d(colour = "black", bins=50, alpha=0.5)+
  xlab("CD3 BUV563")+
  ylab("CD4 APC-Cy7")

# A contour plot of CD3 and CD4 expression over a density plot.
ILC_transf %>%
  ggplot(aes(x = CD3, y = CD4))+        
  geom_bin_2d(bins = 200)+
  scale_fill_gradientn(colors = c("blue",  "yellow", "red"))+
  geom_density_2d(colour = "black")+
  xlab("CD3 BUV563")+
  ylab("CD4 APC-Cy7")

```

<wimr>

# Subsetting data and summary statistics {.tabset .tabset-fade}

While not as efficient as FlowJo, we can subset data after exporting. We wouldn't normally do this in R, but we'll show you how you could subset in R and for the purposes of visualisation.

## Subsetting

```{r}
# Plot a 2-marker plot of CD3 and CD4 expression values. 
ILC_transf %>%
  ggplot(aes(x = CD3, y = CD4))+        
  geom_bin_2d(bins = 200)+
  scale_fill_gradientn(colors = c("blue",  "yellow", "red"))+
  xlab("CD3 BUV563")+
  ylab("CD4 APC-Cy7")+
  geom_vline(xintercept = 1.3)+ #choose the x intercept
  geom_hline(yintercept = 0.6)  #choose the y intercept
```

Subset the dataset based on these thresholds.

```{r}
# This code creates a new column named 'subset' in the 'ILC_transf' dataset.
# The 'subset' column categorises rows based on the values of 'CD3' and 'CD4' columns:
# - "CD3+CD4+": if CD3 > 1.3 and CD4 > 0.6
# - "CD3+CD4-": if CD3 > 1.3 and CD4 <= 0.6
# - "CD3-CD4+": if CD3 <= 1.3 and CD4 > 0.6
# - "CD3-CD4-": if CD3 <= 1.3 and CD4 <= 0.6
# The 'mutate' function from the dplyr package is used to add this new column.

ILC_transf <- ILC_transf %>% 
  mutate(subset = case_when(
    CD3 > 1.3 & CD4 > 0.6 ~ "CD3+CD4+",
    CD3 > 1.3 & CD4 <= 0.6 ~ "CD3+CD4-",
    CD3 <= 1.3 & CD4 > 0.6 ~ "CD3-CD4+",
    TRUE ~ "CD3-CD4-"
  ))
```

<wimr>

### Visualise the subsets in dot plots

```{r}
ILC_transf %>%
  ggplot(aes(x = CD3, y = CD4, color=subset))+        
  geom_point()+
  xlab("CD3 BUV563")+
  ylab("CD4 APC-Cy7")+geom_vline(xintercept = 1.3)+
  geom_hline(yintercept = 0.6)

# Look at different markers coloured by subset
ILC_transf %>%
  ggplot(aes(x = CD56, y = CD16, color=subset))+        
  geom_point()

# Split by subset
ILC_transf %>%
  ggplot(aes(x =CD56, y = CD16))+        
  geom_bin_2d(bins = 200)+
  scale_fill_gradientn(colors = c("blue",  "yellow", "red"))+
  facet_wrap(~subset)

ILC_transf %>%
  ggplot(aes(x =FSC.A, y = SSC.A))+        
  geom_bin_2d(bins = 200)+
  scale_fill_gradientn(colors = c("blue",  "yellow", "red"))+
  facet_wrap(~subset)
```

<wimr>

### Histograms

```{r, warning=FALSE}
ILC_transf %>%
  ggplot(aes(x = CD56, color = subset, fill = subset)) +
  geom_density(alpha = 0.2, linewidth = 1)+
  xlim(quantile(ILC_transf$CD56, 0.001),quantile(ILC_transf$CD56, 0.999))

#using ggridges
ILC_transf %>%
  ggplot(aes(x = CD56, y = subset, fill = subset)) +
  geom_density_ridges(scale = 1.2, alpha = 0.6, colour = "black") +
  theme_ridges() +
  theme(legend.position = "none")+
  labs(y="", x="", title = "CD56")+
  xlim(quantile(ILC_transf$CD56, 0.001),quantile(ILC_transf$CD56, 0.999))
```

You could mass print these using a loop, as we did for the density plots above. You can also adjust the marker you output, as not all markers or combinations of markers are useful.

<wimr>

## Summary statistics

While subsetting is not as efficient as FlowJo, summarising data is!

We can first collect subset percentages and output them in a table, in text, and even in a plot. Then we can calculate summary statistics like MFI & gMFI.

```{r}
# total summary
round(table(ILC_transf$subset)*100/nrow(ILC_transf),2) # % of each subset

# individual marker
ILC_transf %>% 
  group_by(subset) %>% 
  summarise(Percent = round(n()/nrow(ILC_transf)*100,2)) %>%
  mutate(subset = factor(subset, levels = c("CD3+CD4+", "CD3+CD4-", "CD3-CD4+", "CD3-CD4-")))%>%
  mutate(Subsets = "Subsets") %>%
  ggplot(aes(x=Subsets,y = Percent, fill = subset)) +
    geom_bar(stat = "identity", show.legend = FALSE) +
    ylab("Percentage") +
    xlab("") +
    scale_fill_manual(values = c("CD3+CD4+" = "blue", "CD3+CD4-" = "red", "CD3-CD4+" = "green", "CD3-CD4-" = "yellow")) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
    theme(axis.text.y = element_text(size=10)) +
    theme(legend.position = "none") +
    theme(plot.title = element_text(hjust = 0.5, size=10)) +
    ggtitle("Subsets of CD3+/-CD4+/- cells")
```

We can also calculate the percentage of subsets in line, meaning we don't have to change these lines of text based on different results or different data:

-   CD3+CD4+: `r round(sum(ILC_transf$subset == "CD3+CD4+")/nrow(ILC_transf)*100,2)`%

-   CD3+CD4-: `r round(sum(ILC_transf$subset == "CD3+CD4-")/nrow(ILC_transf)*100,2)`%

-   CD3-CD4+: `r round(sum(ILC_transf$subset == "CD3-CD4+")/nrow(ILC_transf)*100,2)`%

-   CD3-CD4-: `r round(sum(ILC_transf$subset == "CD3-CD4-")/nrow(ILC_transf)*100,2)`%

<details>

<summary>**You can also quickly calculate summary statistics like MFI, gMFI, medianFI**:</summary>

<br> **Arithmatic MFI**

Calculate the mean of expression values.

```{r collapse=T}
ILC$subset  <- ILC_transf$subset # Add the subset column to the original data.

ILC %>% 
  group_by(subset) %>% 
  summarise(across(7:31, mean, .names = "MFI_{col}"))

```

**Median FI**
The median.

```{r collapse=T}
ILC %>% 
  group_by(subset) %>% 
  summarise(across(7:31, median, .names = "MFI_{col}"))
```

**Geometric MFI**

The traditional formula written out would be like `ILC_transf %>% prod(CD56)^(1/nrow(CD56))`. However, as our data is *compensated*, there are meaningful negative values and 0s that we don't want to get rid of, which if you run on this data removes two subsets entirely. FlowJo calculates the *graph spaces* as gMFI to enable it to deal with negatives and 0s. Here in this case, it may be better to stick with the arithmetic mean instead of forcing the geometric mean.

</details>

<wimr>

# nxn plots

This code chunk creates a loop to view the nxn plots, and save them to the folder of your choice.

```{r, fig.width=20, fig.height=20, warning=FALSE, eval=F}
# Select only numeric marker columns
marker_cols <- colnames(ILC_transf)[1:25] # adjust if needed

# Make a customised function for plot streamline
custom_bin_plot <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) +
    geom_bin_2d(bins = 50) +
    scale_fill_gradientn(colours = c("blue", "yellow", "red")) +
    theme_minimal()
}

# Generate plots
p <- ggpairs(
  ILC_transf,
  columns = marker_cols,
  lower = list(continuous = custom_bin_plot),
  diag = list(continuous = wrap("densityDiag", alpha = 0.4)),
  upper = list(continuous = wrap("cor", size = 3))
)
ggsave("plots/nxn.png",p, width=25, height=25,unit="in")

rm(p,custom_bin_plot, marker_cols)
```

<wimr>

# Dimension reduction and clustering

There are several ways to perform clustering (e.g. FlowSOM, FuseSOM) and dimension reduction (e.g. PCA, UMAP, tSNE).

In this section, we'll cover **clustering** and **dimension reduction** on our one sample.

## Clustering {.tabset .tabset-fade}

Clustering groups our cells based on the input expressions we choose.

### FlowSOM

[Read more](https://bioconductor.org/packages/release/bioc/html/FlowSOM.html)

```{r, message = FALSE}
# Perform clustering
clusters <- FlowSOM::FlowSOM(
  input=as.matrix(ILC_transf[,1:32]),
  silent = FALSE,
  colsToUse = c(1,4,6:14,16:25),
  nClus = 15)

# Annotate the cluster that the cell belongs to
ILC_transf$flowSOM<- GetMetaclusters(clusters)
#FlowSOM::GetMetaclusterMFIs(clusters)

# Plot CD4 expression values on clusters
ILC_transf %>%
  ggplot(aes(x = CD56, y = flowSOM, fill = flowSOM)) +
  geom_density_ridges(scale = 1.2, alpha = 0.6, colour = "black") +
  theme_ridges() +
  theme(legend.position = "none")+
  labs(y="", x="", title = "CD56")
```

<wimr>

### FuseSOM

[Read more](https://bioconductor.org/packages/devel/bioc/vignettes/FuseSOM/inst/doc/Introduction.html)

```{r, message = FALSE}

risomMarkers <- colnames(ILC_transf)[c(1,4,6:14,16:25)]
clusters <- runFuseSOM(data = as.matrix(ILC_transf[,1:32]), 
                       markers = risomMarkers, 
                       numClusters = 15)

ILC_transf$fuseSOM<- as.numeric(unlist(lapply(str_split(clusters$clusters, "_"), function(x) x[[2]])))

ILC_transf %>%
  ggplot(aes(x = CD56, y = as.factor(fuseSOM), fill=as.factor(fuseSOM))) +
  geom_density_ridges(scale = 1.2, alpha = 0.6, colour = "black") +
  theme_ridges() +
  theme(legend.position = "none")+
  labs(y="", x="", title = "CD56")
```

<wimr>

## Dimension Reduction {.tabset .tabset-fade}

We'll generate a PCA, a tSNE and a UMAP. I'll then run through all of this in the Spectre packge.

```{r message=F, warning=F, eval=T}

pca <- prcomp(as.matrix(ILC_transf[,c(1,4,6:14,16:25)]), center = TRUE)

ILC_transf <- ILC_transf %>% 
  mutate(PC1 = pca$x[,1],
         PC2 = pca$x[,2])

# can take a while
tsne <- Rtsne(as.matrix(ILC_transf[,c(1,4,6:14,16:25)]), dims = 2, perplexity = 30, verbose = TRUE)

ILC_transf <- ILC_transf %>% 
  mutate(tSNE1 = tsne$Y[,1],
         tSNE2 = tsne$Y[,2])

umap <- umap(ILC_transf[,c(1,4,6:14,16:25)])
ILC_transf <- ILC_transf %>% 
  mutate(UMAP1 = umap$layout[,1],
         UMAP2 = umap$layout[,2])

rm(pca,umap,tsne)
```

```{r, eval=F}
saveRDS(ILC_transf,"data/clustered_data.rds")
```

<wimr>

### Visualisations

```{r eval=F}
ILC_transf <- readRDS("data/clustered_data.rds")
```

<details><summary>**Plots:**</summary>

```{r}
ggplot(ILC_transf, aes(PC1,PC2, color= as.factor(flowSOM))) +
  geom_point()

ggplot(ILC_transf, aes(tSNE1,tSNE2, color= as.factor(flowSOM))) +
  geom_point()

ggplot(ILC_transf, aes(UMAP1,UMAP2, color= as.factor(flowSOM))) +
  geom_point()

# Colour overlayed by the expression of one marker.
ggplot(ILC_transf, aes(tSNE1,tSNE2, color= CD3)) +
  geom_point() + 
  scale_color_viridis_c()
ggplot(ILC_transf, aes(tSNE1,tSNE2, color= CD4)) +
  geom_point() + 
  scale_color_viridis_c()
ggplot(ILC_transf, aes(tSNE1,tSNE2, color= CD16)) +
  geom_point() + 
  scale_color_viridis_c()
```

</details>

<wimr>

### Spectre

[Read more](https://immunedynamics.io/spectre/)<br>

Spectre has easy plotting tools to visualize the results of clustering and dimension reduction.

```{r, message = F, warning=FALSE, eval=F}
make.colour.plot(ILC_transf, 
                 "tSNE1", "tSNE2", "subset",  
                 add.label = TRUE,
                 blank.axis = T,
                 path = "plots")

make.colour.plot(ILC_transf, 
                 "tSNE1", "tSNE2", "flowSOM", 
                 col.type = 'factor', 
                 add.label = TRUE,
                 blank.axis = T,
                 path = "plots")
```

<details>

<summary>**Multi-plots:**</summary>

```{r, message = F, warning=FALSE, eval=F}
make.multi.plot(ILC_transf, 
                "tSNE1", "tSNE2", 
                risomMarkers, 
                blank.axis = T, 
                colours = "viridis", 
                path = "plots")
```

</details>

And we can make the heatmap you're used to seeing and save them.

```{r, message = F, warning=FALSE, eval=F}
exp <- do.aggregate(as.data.table(ILC_transf), use.cols = risomMarkers, by = c("flowSOM"))
make.pheatmap(exp, file.name = "plots/Pheatmap_FlowSOM.png", sample.col="flowSOM", plot.cols=risomMarkers)
```

*Spectre has also embedded FlowSOM clustering and dimension reduction tools such as umap in easy codes, such as:*<br>

-   `do.subsample()`
-   `run.flowsom()`
-   `run.umap()` <br>

<wimr>

# Additional Considerations: QC {.tabset .tabset-fade}

When looking at the raw data, or in the middle of data processing, you may consider filtering some cell. In the 1-marker plots below, you can see that there are a few extreme values that are displayed and pushing the majority of values squeezed together. e.g. HLA-DR:

```{r}
ILC_transf %>%
  ggplot(aes(x = HLA.DR))+
  geom_density(linewidth = 0.7, colour = "red", aes(fill = "red"), alpha = 0.5,show.legend = FALSE)+
  ylab("Density")+

# Visualise the values at the top 0.1% and bottom 0.1%.
  geom_vline(xintercept = quantile(ILC_transf$HLA.DR, 0.999), colour = "blue", linewidth = 0.5)+
  geom_vline(xintercept = quantile(ILC_transf$HLA.DR, 0.001), colour = "blue", linewidth = 0.5)+

# We can label the threshold lines: annotate as a text, at the position of the top and bottom 0.01%, and slightly adjust the label position horizontally.
  annotate("text", x = 2, y = 0.5, label = "Top and bottom 0.1%", hjust = -0.1) 
```

<details>

<summary>View them for all channels.</summary>

```{r, message=F}
# Create the empty list
plots_line <- list()

# Create the loop to generate plots.
for (i in 1:25){
  
  # Get the marker names
  Marker <- colnames(ILC_transf)[i]
  
  # Generate the plots and put it into the list
  p <- ILC_transf %>%
  ggplot(aes(x = .data[[Marker]]))+ 
  geom_density(linewidth = 0.7, colour = "red", fill = "red", alpha = 0.3,show.legend = FALSE)+
   xlab(Marker)+
   ylab("Density")+
  
    # Add the newly written lines to visualise the outliers to this loop.
    geom_vline(xintercept = quantile(ILC_transf[[i]], 0.999), colour = "blue", linewidth = 0.5)+
    geom_vline(xintercept = quantile(ILC_transf[[i]], 0.001), colour  = "blue", linewidth = 0.5)
    
  # Put the plots into the list
  plots_line[[length(plots_line)+1]] <- p
}

# Use a grid layout to view the plots
library(ggpubr)
ggpubr::ggarrange(plotlist = plots_line, ncol = 2, nrow = 4)
```

</details>

<br>

To look at this in a more statistical way, you can create a quick data summary of the minimum, maximum, mean, median and the interquartile range.

<details>

<summary>**Expand**</summary>

```{r}
summary(ILC_transf)
```

</details>

Either being artificial or biological, these might interfere how the computer interprets your data in the downstream clustering and dimension reduction. Sometimes you may decide to investigate how these impact your dataset and where they arise from. But if there are plenty of cells, and the % outliers is super small, you might just want to filter them out.<br>

## Filtering by %

You might want to filter out the top and bottom 0.1% of cells . Note that this approach removes the cells that doesn't fit within the top and bottom 0.1% of ANY channel, and removes the cell from the entire dataset. So be cautious about the % so you aren't removing too many cells when they are added up ( we'll use 0.1% all the way for demonstration purposes). <br>

<details>

<summary>**Expand**</summary>

```{r}
# Create an empty list.
outlier_flag <- list()

# Loop through the columns with markers
for (i in 1:25){
  
  # Find the cell values
  value <- ILC_transf[[i]]
  
  # Get the marker names
  Marker <- colnames(ILC_transf)[i]
  
  # Find the thresholds of outliers
  upper <- quantile(value, 0.999, na.rm = TRUE)
  lower <- quantile(value, 0.001, na.rm = TRUE)

  # Flag the outliers as 1, else as 0. And record them into the list.
  flag <- ifelse(value < lower | value > upper, 1, 0)
  outlier_flag[[length(outlier_flag)+1]] <- flag
}
# Turn the flags into a data table object.
outlier_flag <- as.data.table(outlier_flag)
colnames(outlier_flag)<- colnames(ILC_transf)[c(7:31)]
subset <- ILC_transf[rowSums(outlier_flag)==0,]
  
# Calculate the % of cell loss after filtering, then format it as being displayed in less digits, then paste a sentence together.
    paste0("Cell loss: ",format((1-nrow(subset)/nrow(ILC_transf))*100, digits = 3), "%")
```

```{r}

# View the pre- and post-filtered data side by side, putting them in a grid.
cowplot::plot_grid(
  ILC_transf %>%
  ggplot(aes(x = HLA.DR))+
  geom_density(linewidth = 0.7, colour = "red", fill = "red", alpha = 0.3, show.legend = FALSE)+
  ylab("Density"),

subset %>%
  ggplot(aes(x = HLA.DR))+
  geom_density(linewidth = 0.7, colour = "red", fill = "red", alpha = 0.3, show.legend = FALSE)+
  ylab("Density"),

 # number of column equals 1.
 ncol = 1
)

```

</details>

<wimr>

## Filtering by fixed thresholds

You might want to filter out the data that is outside of certain values per channel. A commonly used standard in data cleaning is calculated based on interquartile ranges (IQR). The thresholds are any data points that are 1.5 times of IQR above the 3rd quartile, or 1.5 times of IQR below the 1st quartile. This will keep the majority of data points, however, it might also remove some true variance in the dataset at extreme ends.<br>

<details>

<summary>**Expand**</summary>

```{r}

# Take HLA-DR as an example

# Calculate the interquartile range.
   IQR <- IQR(ILC_transf$HLA.DR)
   Q1 <- quantile(ILC_transf$HLA.DR, 0.25)
   Q3 <- quantile(ILC_transf$HLA.DR, 0.75)
   
# Calculate the upper and lower thresholds
   lowerfence <- Q1 - 1.5*IQR
   upperfence <- Q3 + 1.5*IQR 
 
# Only keep the cells which the HLA-DR value is within the thresholds.   
filt <- ILC_transf[ILC_transf$HLA.DR > lowerfence & ILC_transf$HLA.DR < upperfence, ]

cowplot::plot_grid(
  ILC_transf %>%
  ggplot(aes(x = HLA.DR))+
  geom_density(linewidth = 0.7, colour = "red", fill = "red", alpha = 0.3,show.legend = FALSE)+
  ylab("Density"),
filt %>%
  ggplot(aes(x = HLA.DR))+
  geom_density(linewidth = 0.7, colour = "red", fill = "red", alpha = 0.3,show.legend = FALSE)+
  ylab("Density"),

 # number of column equals 1.
 ncol = 1
)

rm(Q1,Q3,IQR)

```

Or, simply just choose an arbitrary threshold.

```{r}

# Only keep the cells which transformed HLA-DR expression value < 2.
manual <- ILC_transf[ILC_transf$HLA.DR < 2, ]

cowplot::plot_grid(
  ILC_transf %>%
  ggplot(aes(x = HLA.DR))+
  geom_density(linewidth = 0.7, colour = "red", fill = "red", alpha = 0.3,show.legend = FALSE)+
  ylab("Density"),
manual %>%
  ggplot(aes(x = HLA.DR))+
  geom_density(linewidth = 0.7, colour = "red", fill = "red", alpha = 0.3,show.legend = FALSE)+
  ylab("Density"),

 # number of column equals 1.
 ncol = 1
)

```

</details>

# Vote on the next workshop!

<iframe width="640px" height="200px" src="https://forms.office.com/r/TtLWFWysiC?embed=true" frameborder="0" marginwidth="0" marginheight="0" style="border: none; max-width:100%; max-height:100vh" allowfullscreen webkitallowfullscreen mozallowfullscreen msallowfullscreen>

</iframe>
