---
title: "Tidyverse 2: Data Manipulation"
author: "Thomas O'Neil"
date: "2025-02" 
output:
  html_document:
  #workflowr::wflow_html:
    toc: false
---

```{r setup, include=FALSE}
source(url("https://raw.githubusercontent.com/DrThomasOneil/SkillTrees/refs/heads/main/assets/.functions.R")) 
knitr::opts_chunk$set(echo = TRUE, collapse=TRUE, root.dir = dirname(rstudioapi::getActiveDocumentContext()$path))
```


<!---
Notes:
  Introduction
  Data Import/Set up
  Data Wrangling with dplyr 
  - Filtering, selecting, and mutating data.
  - Grouping and Summarising
  - Joining Data Frames
  Reshaping Data with tidyr
<--->

# Introduction

Notes on the tutorial

<div class="hint-goals">
**Goals:** Learn data manipulation with Tidyverse<br>
&nbsp;&nbsp;&#8226; Tidy Data<br> 
&nbsp;&nbsp;&#8226; Select and Mutate data<br> 
&nbsp;&nbsp;&#8226; Group and summarise data<br> 
&nbsp;&nbsp;&#8226; Join two data frames<br> 
&nbsp;&nbsp;&#8226; Pivot data with `tidyr`<br> 

</div>

<read-time><i>**Estimated length**: 1 Hour</i></read-time><wimr>

# Set up{.tabset .tabset-fade}

1. Save this script in a folder called **"202502_Tidyverse2"**

2. Create new folders in your directory.

```{r, eval=F}
dir.create("/plots")
dir.create(paste0(workshop_title, "/data"))
```

3. Download the data.

```{r, eval=F}
# e.g. if its just a csv
download.file(url="https://raw.githubusercontent.com/DrThomasOneil/analysis-user-group/refs/heads/main/docs/r-tutorial/assets/synthetic_data.csv", 
              destfile = "data/data.csv", method='auto', mode='wb')

```

4. 
<wimr>

# Data Wrangling with dplyr{.tabset .tabset-fade}

## Filtering, selecting, and mutating data.

## Grouping and Summarising

## Joining Data Frames

# Reshaping Data with tidyr

# Conclusions

# Acknowledgements

<e>




