---
title: "Data manuipulation and plotting"
author: "Harry Robertson"
date: "2025-02-10"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    embed-resources: true
    css:  
      - https://use.fontawesome.com/releases/v5.0.6/css/all.css
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=T)
```

## Data manipulation and plotting

Data manipulation and plotting are two of the most important tasks in
data analysis. In this tutorial, we will learn how to manipulate data
and create plots using the `dplyr` and `ggplot2` packages in R. But
first, let's learn how to load data into R.

<wimr> 

## Setup{.tabset .tabset-fade}

### Packages

Below is the list of packages that we will be using in this tutorial.
Remember, packages need to be installed only once, but they need to be
loaded every time you start a new R session.

```{r, warning=FALSE, message=FALSE}
# Uncomment the following line to install the package if you haven't already! 
# install.packages(c("dplyr", "ggplot2", "readxl"))

# Load the packages
library(dplyr)
library(ggplot2)
library(readxl)
```
<wimr>

### Create a new project

Set the directory to a common folder you'll continue to use for these workshops. 
From here, you'll be able to create subfolders **in R** for all new workshops. 

```{r, eval=F}
# Set your working directory
knitr::opts_knit$set(root.dir = "/Users/thomasoneil/Desktop/analysis-user-group")

# Create a new folder for todays workshop
dir.create("20250210_Tidyverse")
```
<wimr>

### Download the relevant data

```{r, eval=F}
knitr::opts_knit$set(root.dir = "/Users/thomasoneil/Desktop/analysis-user-group/20250210_Tidyverse")
#Download files
download.file("https://raw.githubusercontent.com/DrThomasOneil/analysis-user-group/refs/heads/main/docs/assets/202502_introSeries/data.rds", "data.rds")
download.file("https://raw.githubusercontent.com/DrThomasOneil/analysis-user-group/refs/heads/main/docs/assets/202502_introSeries/data.csv", "data.csv")
download.file("https://raw.githubusercontent.com/DrThomasOneil/analysis-user-group/refs/heads/main/docs/assets/202502_introSeries/data.tsv", "data.tsv")
download.file("https://raw.githubusercontent.com/DrThomasOneil/analysis-user-group/refs/heads/main/docs/assets/202502_introSeries/data.xlsx", "data.xlsx")
```

If these downloads dont work, run this in your console:

```{r}
set.seed(123) 

data <- data.frame(
  patient_id = 1:1000,
  age = sample(18:90, 1000, replace = TRUE),
  sex = sample(c("Male", "Female"), 1000, replace = TRUE),
  height = rnorm(1000, mean = 170, sd = 10)
)

data$weight <- 0.3 * data$height + rnorm(1000, mean = 19, sd = 5)
data$bmi <- data$weight / (data$height / 100)^2

data$disease_status <- rbinom(1000, 1, prob = plogis(0.05 * (data$age - 50) + 0.1 * (data$bmi - 25)))

data$biomarker <- rnorm(1000, mean = 5, sd = 1) + 2 * data$disease_status

data$on_medication <- rbinom(1000, 1, prob = ifelse(data$disease_status == 1, 0.7, 0.1))
```


<wimr>

## Loading data into R {.tabset}

There are several ways to load data into R. This will largely depend on
the format of the data you are working with. Some of the most common
ways to load data into R are:

-   Loading comma-seperated files (.csv)

-   Loading tab-delimited files (.tsv)

-   Loading Excel files (.xslx)

-   Loading RDS files (R's native file format)


### Loading comma-seperated files (.csv)

One common way is to use the `read.csv()` function to read a CSV file.
For example, to read a file named `data.csv`, you can use the following
code:

```{r, eval = FALSE}
data <- read.csv("data.csv")
head(data)
```
<wimr>

### Loading tab-delimited files (.tsv)

Another way to load data into R is to use the `read.table()` function.
This function can be used to read tab-delimited files, as well as files
with other delimiters. For example, to read a tab-delimited file named
`data.txt`, you can use the following code.

```{r, eval = FALSE}
data <- read.table("data.txt", sep = "\t", header = TRUE) 
# sep = "\t" specifies that the file is tab-delimited. 
# header = TRUE specifies that the first row contains column names. 
head(data)
```

<wimr>

### Loading Excel files (.xslx)

Finally, a lot of collaborators will often give us data in the form of
an Excel file. To read an Excel file into R, you can use the `readxl`
package. First, you need to install the package using the following
code:

```{r, eval = FALSE}
# Uncomment the following line to install the package if you haven't already! 
# install.packages("readxl")
library(readxl)
data <- readxl::read_excel("data.xlsx")
head(data)
```

<wimr>

### Loading RDS files

RDS files are R's native file format. You can save an R object to an RDS
file using the `saveRDS()` function, and then load it back into R using
the `readRDS()` function. For example, to save an object named `data` to
a file named `data.rds`, you can use the following code:

```{r, eval = FALSE}
# Let's first save the data as an RDS file. 
saveRDS(data, "data.rds") 

data <- readRDS("data.rds") # To load the data back into R
head(data)
```

<wimr>

## Simulating data in R

For todays tutorial, we will be simulating our own patient cohort data.
We will create a data frame with 1000 rows and 5 columns. The columns
will be named `patient_id`, `age`, `sex`, `weight`, and `height`.

We will make use of two functions to generate the data: `sample()` and
`rnorm()`.

-   The `sample()` function is used to generate random samples from a
    vector.

-   The `rnorm()` function is used to generate random numbers from a
    normal distribution.

```{r simulating data}
set.seed(123) 

# Simulating dataset
data <- data.frame(
  patient_id = 1:1000,
  age = sample(18:90, 1000, replace = TRUE),
  sex = sample(c("Male", "Female"), 1000, replace = TRUE),
  height = rnorm(1000, mean = 170, sd = 10)
)

# Simulating weight and BMI
data$weight <- 0.3 * data$height + rnorm(1000, mean = 19, sd = 5)
data$bmi <- data$weight / (data$height / 100)^2

# Disease status: More likely in older individuals and those with higher BMI
data$disease_status <- rbinom(1000, 1, prob = plogis(0.05 * (data$age - 50) + 0.1 * (data$bmi - 25)))

# Biomarker: Higher levels in those with disease
data$biomarker <- rnorm(1000, mean = 5, sd = 1) + 2 * data$disease_status

# Medication usage: More common in those with disease
data$on_medication <- rbinom(1000, 1, prob = ifelse(data$disease_status == 1, 0.7, 0.1))

# View first rows
head(data)
```


<wimr>

## Data manipulation with dplyr {.tabset}

The `dplyr` package is a powerful tool for data manipulation in R. It
provides a set of functions that make it easy to filter, arrange, group,
and summarize data. Some of the most commonly used functions in `dplyr`
are: - `filter()`: to filter rows based on a condition - `arrange()`: to
reorder rows - `select()`: to select columns - `mutate()`: to create new
columns.

```{r}
data %>%
  head()
```

### Filtering data

The `filter()` function is used to filter rows based on a condition. For
example, to filter rows where the `age` column is greater than 60, you
can use the following code:

```{r}
older_age_data = data %>%
  filter(age > 60)
head(older_age_data, n = 10)
```

<wimr>

### Arranging data

The `arrange()` function is used to reorder rows based on a variable. For example, to arrange the data by increasing age, you can use the following code:

```{r}
age_increasing_data = data %>%
  arrange(age) 
head(age_increasing_data, n = 10)
```
<wimr>

### Selecting variables

The `select()` function is used to select columns from a data frame. For example, to select only the `sex` and `disease_status` columns, you can use the following code:

```{r}
selected_data <- data %>%
  select(sex, disease_status)
head(selected_data)
```

This is useful if you want to go on to analyze a smaller subset of data.

```{r}
table(selected_data)
```
<wimr>

### Adding variables

The `mutate()` function is used to create new columns based on existing columns. For example, you might want to create a new variable that categorizes patients based on their biomarker level.

```{r}
biomarker_data <- data %>%
  mutate(biomarker_category = ifelse(biomarker > 5, "High", "Low"))
```

<wimr>

### Advanced - Transmute

The `transmute()` function works like `mutate()` but returns only the newly created columns. For example, you might want to create a dataset with only the patient ID and a risk score calculated from age and BMI.

```{r}
risk_data <- data %>%
  transmute(
    patient_id,
    risk_score = 0.05 * age + 0.1 * bmi
  )
head(risk_data)
```
<wimr>

### Advanced - case_when

The `case_when()` function is a more flexible version of `ifelse()`. It allows you to specify multiple conditions and corresponding values. For example, you might want to create a new variable that categorizes patients based on their BMI.

```{r}
data <- data %>%
  mutate(bmi_category = case_when(
    bmi < 18.5 ~ "Underweight",
    bmi >= 18.5 & bmi < 25 ~ "Normal",
    bmi >= 25 & bmi < 30 ~ "Overweight",
    bmi >= 30 ~ "Obese"
  ))
head(data, n = 10)
```

<wimr>

## Plotting with ggplot2 {.tabset}

The `ggplot2` package is a powerful tool for creating plots in R. It
provides a flexible and intuitive syntax for creating a wide variety of
plots, including scatter plots, bar plots, line plots, and more.
 
The basic idea of `ggplot2` is to create a plot in layers. You start by creating a new plot using the `ggplot()` function, and then add layers to the plot using functions like `geom_point()`, `geom_line()`, `geom_bar()`, and so on.

Below we will initiate our first plot!

```{r}
library(ggplot2)
ggplot(data, aes(x = weight, y = height))
```

Obviously this isn't useful yet, but we will now add layers on top of our canvas to create a plot!

### Scatter plots

The `ggplot()` function is used to create a new plot, and the `geom_point()` function is used to add points to the plot. 
```{r}
ggplot(data, aes(x = weight, y = height)) + geom_point()
```

We can also customize the plot by changing the color and transparency of the points within the `geom_point()` function!

```{r}
ggplot(data, aes(x = weight, y = height)) + geom_point(color = "green4", alpha = 0.5)
```

We can also add titles and labels to the plot using the `labs()` function.

```{r}
ggplot(data, aes(x = weight, y = height)) + geom_point(color = "green4", alpha = 0.5) +
  labs(title = "Scatter Plot of Weight vs Height",
       x = "Weight (kg)",
       y = "Height (cm)")
```

We can also change the theme of the plot using the `theme_bw()` function! 

```{r}
ggplot(data, aes(x = weight, y = height)) + geom_point(color = "green4", alpha = 0.5) +
  labs(title = "Scatter Plot of Weight vs Height",
       x = "Weight (kg)",
       y = "Height (cm)") +
  theme_bw()
```

<wimr>

### Box plots

```{r}
ggplot(data, aes(x = factor(disease_status), y = bmi, fill = factor(disease_status))) + geom_boxplot()
```

The current set up of using 0 and 1 to represent disease status is not very informative. Let's change the labels to "No Disease" and "Disease" using the `factor()` function. We will make use of `mutate()` to create a new column with the new labels.

```{r}
box_plot_data <- data %>%
  select(disease_status, bmi) %>%
  mutate(disease_status = factor(disease_status, levels = c(0, 1), labels = c("No Disease", "Disease")))
ggplot(box_plot_data, aes(x = disease_status, y = bmi, fill = disease_status)) + geom_boxplot() 
```

<wimr>

### Line plots

For illustrative purposes, you can simulate a trend over age by summarizing the average biomarker level per age group. Here we use two new functions `group_by()` and `summarize()` from the `dplyr` package to calculate the average biomarker level per age group.   
Simply, the `group_by()` function is used to group the data by a variable, and the `summarize()` function is used to calculate summary statistics for each group.

```{r}
age_trend <- data %>%
  group_by(age) %>%
  summarize(avg_biomarker = mean(biomarker))

ggplot(age_trend, aes(x = age, y = avg_biomarker)) +
  geom_line(color = "purple", size = 1) +
  labs(title = "Average Biomarker Level by Age",
       x = "Age",
       y = "Average Biomarker Level") +
  theme_bw()
```

<wimr>

### Bar plots

```{r}
data %>%
  mutate(bmi_category = case_when(
    bmi < 18.5 ~ "Underweight",
    bmi >= 18.5 & bmi < 25 ~ "Normal",
    bmi >= 25 & bmi < 30 ~ "Overweight",
    bmi >= 30 ~ "Obese"
  )) %>%
ggplot(aes(x = bmi_category, fill = factor(on_medication))) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("lightgrey", "dodgerblue"),
                    labels = c("Not on Medication", "On Medication")) +
  labs(title = "Proportion of Medication Use by BMI Category",
       x = "\nBMI Category",
       y = "Proportion",
       fill = "Medication") +
  theme_minimal()
```

<wimr>

### Histograms

```{r}
ggplot(data, aes(x = biomarker)) +
  geom_histogram(binwidth = 0.5, fill = "cornflowerblue", color = "black") +
  labs(title = "Histogram of Biomarker Levels",
       x = "Biomarker Level",
       y = "Frequency") +
  theme_minimal()
```

<wimr>

## Advanced plotting {.tabset}

`ggplot2` has a lot of advanced features that allow you to create more complex and customized plots. 

### Plotting multiple layers

You can plot multiple datasets on the same plot by adding multiple `geom_` layers. For example, let’s plot the average biomarker level by age group and overlay the individual biomarker levels. (This doesn't make a ton of sense but just shows you how to overlay multiple layers!)

```{r}
ggplot(age_trend, aes(x = age, y = avg_biomarker)) +
  geom_line(color = "purple", size = 1) +
  geom_point(data = data, aes(x = age, y = biomarker), color = "steelblue", alpha = 0.5) +
  labs(title = "Average Biomarker Level by Age with Individual Data Points",
       x = "Age",
       y = "Biomarker Level") +
  theme_bw()
```

<wimr>

### Plotting with facets

Faceting is a powerful technique for creating multiple plots based on the levels of a categorical variable. For instance, we can compare biomarker distributions between patient sexes.

```{r}
ggplot(data, aes(x = biomarker, fill = factor(disease_status))) +
  geom_histogram(binwidth = 0.5, color = "black", alpha = 0.7) +
  facet_wrap(.~ sex) +
  scale_fill_manual(values = c("lightblue", "tomato"),
                    labels = c("No Disease", "Disease")) +
  labs(title = "Biomarker Distribution by Disease Status and Sex",
       x = "Biomarker Level",
       y = "Count",
       fill = "Disease Status") +
  theme_bw()
```

<wimr>

### Plotting with statistics

You can overlay statistical summaries (such as regression lines) using `geom_smooth()`. For example, let’s add a regression line to the scatter plot of weight versus height.

```{r}
ggplot(data, aes(x = weight, y = height)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  labs(title = "Scatter Plot of Weight vs Height with Regression Line",
       x = "Weight (kg)",
       y = "Height (cm)") +
  theme_minimal()
```

We can also perform basic statistical tests between groups. For example, we can compare the average BMI between patients with and without disease using the `stat_compare_means()` function from the `ggpubr` package.

```{r}
# Uncomment the following line to install the package if you haven't already!
# install.packages("ggpubr")
library(ggpubr)
box_plot_data <- data %>%
  select(disease_status, bmi) %>%
  mutate(disease_status = factor(disease_status, levels = c(0, 1), labels = c("No Disease", "Disease")))
ggplot(box_plot_data, aes(x = disease_status, y = bmi, fill = disease_status)) + geom_boxplot() + theme_bw() + stat_compare_means(method = "t.test")
```

<wimr>

## Conclusion

In this tutorial, we learned how to load data into R, manipulate data
using the `dplyr` package, and create plots using the `ggplot2` package.
We also learned how to customize plots and create more advanced plots.
These skills are essential for any data analyst or scientist working
with R.
