---
title: ""
output:
  workflowr::wflow_html:
    toc: no
---
<div style = "text-align: center">

<span style="font-size:2em; color: #DC582A;"><i>**Spectral Unmixing in R**</i></span>   

#### **August 12th**

*On your own device, for <strong>FREE!</strong>*

Spectral cytometry data and spectral unmixing can be daunting to look at first,  
but it is simply linear algebra. Yuchen promises you it is not as hard as you might  
think, and that you can absolutely do it on your own.  

**Lecture** (11am - 11:30am); **Workshop** (12pm - 2pm)

<span style = "color:red; font-size:1.7em"><i>**Lunch from 11:30!**</i></span>

**Register [here](https://forms.cloud.microsoft/r/HtM8qDXQhi) or below, for catering purposes!**

[Zoom link](https://uni-sydney.zoom.us/j/84164373602)

<hr></div>

# {.tabset .tabset-fade}

<div>

## Event Details {.tabset .tabset-fade}

### **Lecture**: 11am - 11:30pm (WIMR Seminar Room & Zoom)

**What to expect in the lecture:**

- Learn about spectral flow data structure and the principle of unmixing

### **Workshop**: 12pm - 2pm (WIMR Seminar Room & Zoom)

**What to expect in the workshop:**
 
- Set up your R environment for spectral flow.  
- Perform spectral unmixing in R.  
- Inspect and gate unmixed data in R.  
- Extract summary statistics from gated data.  

## Setup for workshop {.tabset .tabset-fade}

### Requirements

An `R` version of at least 4.5.x and `Rstudio.`

`Rtools` 4.5 or 4.6 depending on the R version.

Necessary R packages: `biocManager`, `flowCore`, `flowSpectrum`, `ggcyto`, `matrixStats`, `data.table`.

---

### Install instructions

To download `biocManager`:

```{r eval=F}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(version = "3.23") #or change the version if you use earlier R versions.
```

`biocManager` is required to download flowCore and ggcyto and their dependencies.

```{r eval=F}
BiocManager::install(c("ggcyto","flowCore"))
```


To download `flowSpectrum` through GitHub:

```{r eval=F}
devtools::install_github('hally166/flowSpectrum')
```


If required, `matrixStats` and `data.table` can be downloaded using the standard CRAN install argument `install.packages("package name")`

---

### Download the Data

The repository and example data that we will use can be accessed through this [link](https://filesender.aarnet.edu.au/?s=download&token=96c074a6-275a-4f44-b946-d749b4fb0336). We will work through the **Workshop.zip**. The backup is also downloadable which contains unmixed data in case the users struggle on the day. 

---

#

<span style="display:inline-block; width:80%;"><wimr></span>

<span style = "color:red;font-size:1.5em">**Sign up below:**</span>

<iframe width="640px" height="480px" src="https://forms.cloud.microsoft/r/HtM8qDXQhi?embed=true" frameborder="0" marginwidth="0" marginheight="0" style="border: none; max-width:100%; max-height:100vh" allowfullscreen webkitallowfullscreen mozallowfullscreen msallowfullscreen> </iframe>
